---
title: "BPIC-2015"
author: "lars pijnappel"
date: "30 augustus 2015"
output: html_document
---

***
More information regarding these BPI Challenge datasets can be found on the site of the [11th International Workshop on Business Process Intelligence 2015](http://www.win.tue.nl/bpi/2015/challenge).

***


## Import data by means of `lapply` and `ldply`
> and show imported data size (MB)

```{r}
fpn <- list.files( 'data BPIC-2015', pattern = '.csv', full.names = TRUE ) ## file path name

fn <- basename( fpn )   ## file name
names( fpn ) <- basename( fpn )

## total size of 5 csv files = 74,30 MB
```

### lapply
```{r}
## lapply is fast and output is stored in a list object.
# > system.time( lapply( fpn, read.csv ))
#    user  system elapsed 
#    7.63    0.06    7.69 

l_csv <- lapply( fpn, read.csv ) ## import selected csv files
comment( l_csv ) <- "data imported with lapply"
paste( as.integer( object.size( l_csv )/1024^2 ), " MB", sep="" )
```

### ldply
```{r}
library( plyr )
## ldply is faster and output is stored in a single data.frame
# > system.time( ldply( fpn , read.csv, stringsAsFactors = FALSE ))
#    user  system elapsed 
#    5.66    0.06    5.80

df_csv <- ldply( fpn, read.csv, stringsAsFactors = FALSE )
comment( df_csv ) <- "data imported with ldply + read.csv (default arguments)"
paste( as.integer( object.size( df_csv )/1024^2 ), " MB", sep="" )

# df_csvchar <- ldply( fpn, .fun = read.csv, colClasses = "character", stringsAsFactors = FALSE )
# comment( df_csvchar ) <- "data imported with ldply + read.csv (colClasses='character')"
# paste( as.integer( object.size( df_csvchar )/1024^2 ), " MB", sep="" )
```


## Show numbers of (1) observations and (2) variables per file

### lapply
```{r}
l_dim <- lapply( l_csv, dim )
names( l_dim ) <- fn
l_dim
```

### ldply
```{r}
## ldply has added variable .id (which contains the source file name). 
## Consider this variable as categorical and trim the filename to first 8 characters.
df_csv$.id <- factor( strtrim( df_csv$.id, 8 ))      
as.data.frame( summary( df_csv$.id ) )
dim( df_csv )

# df_csvchar$.id <- factor(df_csvchar$.id)      ## Consider this variable as categorical
# as.data.frame( summary( df_csvchar$.id ) )
# dim( df_csvchar )
```


## Determine if all files have same variable names 
> check consistency column names and show them

### lapply
```{r}
cn <- lapply( l_csv, colnames )
for ( i in seq_along(cn[-length(cn)]) ) { ## loop, but exclude last cn
  x <- identical( cn[[i]], cn[[i+1]] )
  print( paste('Files ', i, ' + ', i+1, ' with identical colnames: ', x, sep = '') )
}

# cn[1]
```

### ldply
```{r}
names( df_csv )
```


## Show number of observations per file
### lapply
```{r}
barplot( unlist(lapply( l_csv, nrow )), ylab = 'observations', col = 'cyan' )
```

### ldply
```{r}
barplot( summary( df_csv$.id ), ylab = 'observations', horiz = TRUE, col = 'green')
```


## Show the structure of the imported data
### ldply
```{r}
str( df_csv )
# str( df_csvchar )
```


## Show a summary of the the imported data
> BEFORE factorizing several variables

```{r}
summary( df_csv )
# summary( df_csvchar )
```


## Show a summary of the the imported data
> AFTER factorizing several variables
```{r eval=FALSE}
df2 <- df_csv
df2$Resource <- factor( df2$Resource )
df2$X.case..IDofConceptCase <- factor( df2$X.case..IDofConceptCase )
df2$X.case..Responsible_actor <- factor( df2$X.case..Responsible_actor )
df2$X.case..caseProcedure <- factor( df2$X.case..caseProcedure )
df2$X.case..caseStatus <- factor( df2$X.case..caseStatus )
summary( df2$X.case..caseStatus )
summary( df2 )
```


## Show content of first obs. in readable format
### lapply
```{r eval=FALSE, echo=FALSE}
t(head( l_csv[[1]], 1 ))
```

### ldply
```{r}
t(head( df_csv, 1 ))
# t(head( df_csvchar, 1 ))
```


## Count observations for each resource
> variables: Resource / monitoringResource / Responsible_actor

### lapply
```{r eval=FALSE echo=FALSE}
## first create a working df
csv1 <- l_csv[[1]]
## variable 'responsible actor' is categorical, not quantitative.
csv1$X.case..Responsible_actor <- factor( csv1$X.case..Responsible_actor)
csv2 <- csv1$X.case..Responsible_actor
comment(csv2) <- "one way of counting observations per responsible actor"
# summary(csv1$X.case..Responsible_actor)
csv3 <- as.matrix( summary(csv2) )
csv3
```

### ldply

Defining function `role_summary` to summarize the total events per role per organization. By default, only cross organizational roles are returned. 

```{r echo=FALSE, eval=FALSE}
##### Defining functions `role_xorg` and `role_summary` to summarize the total events per role per organization.

# df <- df_csv
# role <- "Resource"
# role <- "X.case..Responsible_actor"
# role <- "monitoringResource"

role_summary <- function ( role ) {
        ## GOAL: sums the total events (observations) per role per organization (csv). 
        ## result is sorted first by role (row=ascending) then by column
     
        ## convert input parameters into an R object dfv (df$variable)
        ## www.r-bloggers.com/converting-a-string-to-a-variable-name-on-the-fly-and-vice-versa-in-r/
        dfv <- eval( parse( text = paste("df_csv", "$", role, sep = "" )))
        ## events count per role, split by csv file
        rce <- tapply( factor( dfv ), df_csv$.id, summary )
        summary( rce )
        
        ## QAD: only csv 1 contains NA's for responsible actors: remove it.
        if ( role == "X.case..Responsible_actor" ) rce[[1]] <- rce[[1]][ -length( rce[[1]] ) ]
        
        ## convert the list object rce into a matrix object, then sort and return the results
        rce <- do.call( cbind, rce )
        rce[ order( rce[,1], rce[,2], rce[,3], rce[,4], rce[,5], decreasing = TRUE ), ]   
}

role_xorg <- function ( role ) {
        ## GOAL: return roles which have events (observations) in more than 1 organization (csv file)
        
        ## first collect a summary for all roles for all organizations
        rs <- role_summary( role )
        
        ## count for each role the number of organizations where events have been observed
        ## if count is > 1 then a cross organizational role has been detected.
        xo <- nrow( rs )
        for ( i in seq_len( nrow( rs ))) { 
                xo[i] <- sum( c( rs[i,1] > 0, rs[i,2] > 0, rs[i,3] > 0, rs[i,4] > 0, rs[i,5] > 0 ))
        }
        rs[ which( xo > 1 ), ]     
}
```

```{r}
role_summary <- function ( role, xorg = TRUE ) {
        ## GOAL: sums the total events (observations) per role per organization (csv). 
        ## result is sorted first by role (row=ascending) then by column
        ## when xorg = TRUE then only cross organizational roles are returned (otherwise all roles)
        
        ## convert input argument role into an R object dfv (df$variable)
        ## www.r-bloggers.com/converting-a-string-to-a-variable-name-on-the-fly-and-vice-versa-in-r/
        dfv <- eval( parse( text = paste("df_csv", "$", role, sep = "" )))
        ## for each role count events. Then split by csv file
        rce <- tapply( factor( dfv ), df_csv$.id, summary )
        summary( rce )
        
        ## QAD: only csv 1 contains NA's for responsible actors: remove it.
        if ( role == "X.case..Responsible_actor" ) rce[[1]] <- rce[[1]][ -length( rce[[1]] ) ]
        
        ## convert the list object rce into a matrix object, then sort the results
        rce <- do.call( cbind, rce )
        rce <- rce[ order( rce[,1], rce[,2], rce[,3], rce[,4], rce[,5], decreasing = TRUE ), ]   
        
        if ( xorg == TRUE ) {
                ## count for each role the number of organizations where events have been observed
                ## if count is > 1 then a cross organizational role has been detected.
                xo <- nrow( rce )
                for ( i in seq_len( nrow( rce ))) { 
                        xo[i] <- sum(c( rce[i,1] > 0, rce[i,2] > 0, rce[i,3] > 0, rce[i,4] > 0, rce[i,5] > 0))
                }
                rce <- rce[ which( xo > 1 ), ]  
        }
        
        rce
}
```

### Cross Organisational Roles

```{r eval=FALSE, echo=FALSE}
role_xorg ( "Resource" )
role_xorg ( "monitoringResource" )
## When nr. of returned results = 1 then rownames is not shown..
role_xorg ( "X.case..Responsible_actor" )
```


#### Resource
Retrieve all `Resource` roles who act cross organizational.
```{r}
role_summary( "Resource", TRUE )
```

#### monitoringResource
Retrieve all `monitoringResource` roles who act cross organizational.
```{r}
role_summary( role = "monitoringResource" )
```

#### Responsible_actor
First retrieve only `Responsible_actor` roles who act cross organizational. Then list all `Responsible_actor` roles.
```{r}
role_summary( role = "X.case..Responsible_actor" )
role_summary( role = "X.case..Responsible_actor", FALSE )
```


## Per csv, indicate nr. of unique values

### 1
```{r eval = FALSE, echo = FALSE}
l <- data.frame( matrix( nrow=5, ncol=7 ))
comment(l) <- "per csv, indicate nr. of unique values"
colnames(l) <- c("caseStatus", "caseProcedure", "IDofCOnceptCase", "concept.name", "ActivityNameNL", "ActivityNameEN", "requestComplete")
for ( i in seq_along( l_csv )) {
        l[i, 1] <- paste( levels( l_csv[[i]][,"X.case..caseStatus"] ), collapse = "/" )
        l[i, 2] <- paste( levels( l_csv[[i]][,'X.case..caseProcedure'] ), collapse = "/" )
        l[i, 3] <- length( unique( na.omit( l_csv[[i]][,'X.case..IDofConceptCase'] )))   # Remove NA's
        l[i, 4] <- length( levels( l_csv[[i]][,'concept.name'] ))
        l[i, 5] <- length( levels( l_csv[[i]][,'activityNameNL'] ))
        l[i, 6] <- length( levels( l_csv[[i]][,'activityNameEN'] ))
        l[i, 7] <- length( unique( l_csv[[i]][,'X.case..requestComplete'] ))
}
comment( l )
l
```

### 2
```{r eval = FALSE, echo = FALSE}
anNL <- data.matrix( unique(l_csv[[1]][,'activityNameNL'] ))
anNL <- cbind( anNL, 1:nrow( anNL ))
anEN <- data.matrix( unique( l_csv[[1]][,'activityNameEN'] ))
anEN <- cbind( anEN, 1:nrow( anEN ))
cor( as.numeric( anNL[,2] ), as.numeric( anEN[,2] ))

# correlation acitivity names
can <- cbind( l_csv[[1]][,'activityNameNL'], l_csv[[1]][,'activityNameEN'])
tail(can)
cor( can[,1], can[,2] )
```


## Activities EN
```{r eval = FALSE, echo = FALSE}
anEN
```


## Identicals
```{r eval = FALSE, echo = TRUE}
identical( l_csv[[1]][1,'action_code'], l_csv[[1]][1,'concept.name'])
d1 <- as.character( l_csv[[1]][,'action_code'] )
d2 <- as.character( l_csv[[1]][,'concept.name'] )
identical( d1, d2)
d3 <- as.character( l_csv[[1]][,'Activity'] )
d4 <- as.character( l_csv[[1]][,'lifecycle.transition'] )

d1[which(d2 %in% d1) == TRUE]
d2[which(d2 %in% d1) == TRUE]
d1[ ( d1 == d2 ) == FALSE]
d2[ ( d1 == d2 ) == FALSE]

d_all <- as.data.frame( cbind(d1,d2,d3,d4) )
str(d_all)
```
